Local Sensitivity Analysis of a Supercritical Extraction Model

### 2025-03-30 Article_Sensitivity_Analsysis-Chamomile.tex
30.03

	{\color{blue}A key novelty of this work lies in its rigorous application of derivative-based local sensitivity analysis to a first-principles model of supercritical fluid extraction, specifically targeting the extraction of chamomile oil. While numerous studies have developed empirical or semi-empirical models for supercritical extraction, many of these either omit local sensitivity analyses altogether or provide only basic parametric comparisons (\citet{Santos2000}, \citet{Fiori_2007} or \citet{Zahedi2010}). In contrast, the present approach integrates a quasi-1D partial-differential-equation model with automatic differentiation of the mass and energy balances, enabling a detailed exploration of how infinitesimal changes in operating parameters propagate through the system to influence solute concentrations and overall extraction yield. Although this work focus on the influence of pressure, the same methodology can be analysed any other parameter of control. Notably, this level of detail is crucial near the critical region of CO$_2$, where small deviations in process conditions can trigger large changes in solvent properties (such as density and solubility). This approach thus provides a systematic and quantitative measure of the system’s sensitivity to parameter fluctuations across different extraction regimes, from kinetic- to diffusion-controlled.}

@@

	{\footnotesize
		\begin{equation}
			\label{Model_kinetic}
			r_e = -\cfrac{D_i^R \gamma }{ \mu l^2 } \left( c_s  - \cfrac{\rho_s c_f }{ k_m \rho_f }  \right)
	\end{equation} }

@@

	%\subfile{Sections/Results_Sensitivity}
	\section{Results} \label{CH: Results}
	The details of the process model and parameters are provided in \citet{Sliczniuk2024}. The process model was validated in the following range of operating conditions: temperatures between $30 - 40 ^\circ C$, pressures between $100 - 200$ bar, and mass flow rates between $3.33-6.67 \cdot 10^{-5}$ kg/s. This study examines the impact of parameter change on the state space and extraction yield through local sensitivity analysis. The local sensitivity analysis results can be interpreted as the system's response to an infinitesimal deviation in parameters. {\color{blue}This work focus on two selected cases, but the same methodology can be extended to any other model parameter. The first case corresponds to the analysis of empirical correlations on the process model. The zero'th order term of the multiple linear regression for $\Upsilon$ (denoted as $\Upsilon(0)$) was selected. The second case corresponds to the investigation of the operating condition on the state-space. Particularly, on the influence of pressure variations on the model output. The first analysis is performed at the midpoint of the validated range: $35 ^\circ C$, 150 bar and 5 $\cdot 10^{-5}$ kg/s. The second analysis is split into two steps. First, the influence of the pressure change on the state-variables (concentration of the solute in the sold and fluid phases) are considered at the midpoint of the validated range. Secondly, the yield analysis is conducted at the same temperature and mass flow-rate conditions but at different values of pressure (100, 125, 150, 175 and 200 bar).}

@@

	{\color{blue}The $\Upsilon$ term describes the curvature of decay function $\gamma$, and can be associated with uneven distribution of the solute in solid particles. As the zero'th order term is not directly affect by independent variables (as it represents the intercept), it physical interpretation becomes straight forward. The increment $\Upsilon(0)$ makes the $\gamma$ steeper, which means that the $D_i^R$ is multiplied by a smaller number, hence it is expected to observe slower extraction rate.

@@

	Figure \ref{fig:Sensitivty_Gamma_CS} illustrates the sensitivity of the solute concentration in the solid phase to change into zero'th order term in the equation for $\Upsilon$. As discussed by \citet{Sliczniuk2024}, the system is far from the saturation point, which indicates that the concentration gradient is controlled mainly by $c_s$. Consequently, the extraction kinetic term act as independent on $c_f$, which explain a uniform response in solid phase visible in Figure \ref{fig:Sensitivty_Gamma_CS}. The positive value of $\frac{dc_s}{d\Upsilon(0)}$ can be interpreted as an slower loss of the solute from particles and is observed when the extraction slows down. Due to lower reaction rate more solute stays in the particles, but as the time approach infinity the solute will be washed out from the solid matrix. Consequently, it is expected to observed $\frac{dc_s}{d\Upsilon(0)}$ reaching a maximum and asymptotically diminish to zero.

@@

	Figure \ref{fig:Sensitivty_Gamma_CF} illustrates the sensitivity of the solute concentration in the fluid phase with respect to change into zero'th order term in the equation for $\Upsilon$. As previously discussed, the extraction rate slowed down due to change of $\Upsilon(0)$ uniformly along the spatial domain. Consequently, the amount of delivered to the fluid phase is reduced by the same quantity along the spatial domain. That leads to decrease of the solute concentration $c_f$ in the fluid phase, which is characterized by negative sign of the $\frac{dc_f}{d\Upsilon(0)}$. The spatial distribution of $\frac{dc_f}{d\Upsilon(0)}$ in the figure is non-uniform, reflecting the effects of advection. Due to mass conservation, the remaining solute will reach the solid-fluid interface slower hence it is expected to observe a little positive values of $\frac{dc_f}{d\Upsilon(0)}$ in the later stage of extraction. Eventually it is expected to see sensitivities reaching zero asymptotically.

@@

	Figure \ref{fig:Sensitivty_Gamma_Y} show the normalized sensitivity curve of with respect to the yield. Due to slower extraction rate, the sensitivity curve is characterized by negative sign. The drop of the sensitivity curve can be interpreted as less solute coming out of the extractor if compared with the reference case. Then the sensitivity curve reach minimum, which can be also visible in Figure \ref{fig:Sensitivty_Gamma_CF}. Eventually, the remaining solute is released to the fluid phase and the sensitivity curve reach asymptotically zero. When the sensitivity function of a dynamic system approaches zero as time goes to infinity and remains at that value, it indicates that the system's long-term behavior becomes independent of the parameter in question. The parameter has only influence for the system's transient dynamics, it has no effect on the steady-state solution. This means that after the system has settled, variations in that parameter do not affect the final output. In terms of SFE, it means that all the solute has been removed from the system.}

@@

	{\color{blue}The motivation behind perform sensitivity analysis model validation and evaluation of parameters effect on the model output. By analysis the influence of parameters potential modelling errors and unfeasible behaviours can be identified. This is particularly useful technique, when introducing an empirical correlation. In this work, the influence of the intersect from the empirical correlation for $\Upsilon$ was tested. As expected, the analysis showed that increment of $\Upsilon(0)$ slows down the extraction rate, leading to negative values of $\frac{dc_s}{d\Upsilon(0)}$ and $\frac{dc_s}{d\Upsilon(0)}$, and alternating values of $\frac{dc_f}{d\Upsilon(0)}$. The physical interpretation of the $\Upsilon(0)$ match obtained results. The effect of the parameter change can be summarized by extreme and final values of the normalized sensitivity of the output (0.47 and 0, respectively).}

	pressure. Under the selected operating conditions, it was observed for all the cases that a pressure increment enhances mass transfer, leading to a faster mass transfer of solute from the particles. Consequently, negative sensitivities can be observed in the diagram for the solid phase. The corresponding response in the fluid phase is characterized by positive-valued sensitivities, indicating that a greater amount of solute is transported into the fluid phase. As a result, the extraction yield is improved and characterized by positive sensitivities. Given that the results of the local sensitivity analysis depend on the operating point, the analysis was repeated at different pressures. The normalized sensitivity coefficients provide a basis for generalizing the discussion of the results and comparing the influence of various parameters. The normalized sensitivity results were found to vary within the range of -0.17 to 0.6. The overall effect of each parameter can be evaluated by integrating the sensitivity curves. Similar to the analysis of yield sensitivity, the integration was performed over extended simulation time (until all sensitivities approached zero). The values of cumulative integral are presented in Table \ref{table:Cumulative_integral}. The positive sign of these values indicates that increased pressure enhances the extraction yield under all investigated operating conditions. The order of magnitude of the integrals allows to identify at which operating condtions the pressure increase is the most influential.

@@

	{\color{blue}The findings of this work are particularly valuable for industrial-scale supercritical extraction processes, where achieving both operational efficiency and product quality can be challenging. The local sensitivity analysis demonstrates that even small adjustments in operating parameters can yield disproportionately large improvements in extraction rates, especially near the fluid’s critical point. By quantifying how parameter variations affect the model states, the high-efficiency operating setpoints can be identified. Moreover, linking local sensitivity analysis to numerical optimization establishes a systematic framework for industrial process design and scale-up. Rather than relying heavily on trial-and-error or resource-intensive pilot studies, the model-based approach can be used to pre-identify the region of interests. This approach reduces both the time and material costs needed to reach profitable extraction rates, moving industrial operations toward cost-effective production.}

	It can be concluded that the system responses at low pressures were stronger than those at higher pressures. This behaviour can be explained by the rapid changes in solvent properties occurring near the critical point of CO$_2$.

@@

	% ===================================================
	% Bibliography
	% ===================================================
	%% Loading bibliography style file
	%\clearpage
	%\newpage
	%\bibliographystyle{model1-num-names}
	\bibliographystyle{unsrtnat}
	\bibliography{mybibfile}

>>>>

	{\color{blue}A key novelty of this work lies in its rigorous application of derivative-based local sensitivity analysis to a first-principles model of supercritical fluid extraction. While multiple publications available in the literature discuss application of sensitivity analysis on a supercritical extraction model, many of these either omit local sensitivity analyses and focus on basic parametric comparisons (\citet{Santos2000}, \citet{Fiori_2007} or \citet{Zahedi2010}). In contrast, the present approach integrates a quasi-1D partial-differential-equation model with automatic differentiation of the mass and energy balances, enabling a detailed exploration of how infinitesimal changes in parameters, which propagates propagate through the system. As the result, corresponding system response of the state space (with emphasis on solutes concentrations on solid and liquid phase as well as yield). Although this work focus on the influence of a few selected parameters, the same methodology can be extended to any other parameter. Notably, this level of detail is crucial near the critical region of CO$_2$, where small deviations in process conditions can trigger large changes in solvent properties (such as density and solubility). This approach thus provides a systematic and quantitative measure of the system’s sensitivity to parameter fluctuations across different extraction regimes, from kinetic- to diffusion-controlled.}

@@

	{\footnotesize
		\begin{equation}
			\label{Model_kinetic}
			r_e = \cfrac{D_i^R \gamma }{ \mu l^2 } \left( c_s  - \cfrac{\rho_s c_f }{ k_m \rho_f }  \right)
	\end{equation} }

@@

	%\subfile{Sections/Results_Sensitivity}
	\section{Results} \label{CH: Results}
	Details of the process model and its parameters can be found in \citet{Sliczniuk2024}. The model was validated under the following operating conditions: temperatures from $30$ to $40 ^\circ C$, pressures from $100$ to $200$ bar, and mass flow rates from $3.33$ to $6.67 \cdot 10^{-5}$ kg/s. In this study, the local sensitivity analysis is used to investigate how changes in model parameters affect the state space and extraction yield, where the results reflect the system’s response to infinitesimal deviations in parameter values. Although this work focuses on two specific cases, the methodology can be extended to any model parameter. The first case assesses the empirical correlations embedded in the process model, specifically the zeroth-order term $\Upsilon(0)$ in the multiple linear regression for $\Upsilon$. The second case examines the effect of operating conditions, particularly pressure, on the state space and model outputs. For the first analysis, the midpoint of the validated range ($35 ^\circ C$, $150$ bar and $5\cdot 10^{-5}$) was selected. The second analysis is carried out in two steps. First, we examine the impact of pressure changes on the state variables (solute concentrations in both solid and fluid phases) at the same midpoint conditions. Second, we perform a yield analysis at the same temperature and mass flow rate, but at different pressures (100, 125, 150, 175, and 200 bar).

@@

	{\color{blue}The $\Upsilon$ term characterizes the curvature of the decay function $\gamma$ and can be associated with the uneven distribution of solute within solid particles. Because the zeroth-order term ($\Upsilon(0)$) corresponds to an intercept rather than depending on independent variables, its physical interpretation is straightforward. Increasing $\Upsilon(0)$ steepens $\gamma$, which means $D_i^R$ is multiplied by a smaller factor and the extraction rate consequently slows down.

@@

	Figure \ref{fig:Sensitivty_Gamma_CS} illustrates the sensitivity of the solute concentration in the solid phase to change into zero'th order term in the equation for $\Upsilon$. As discussed by \citet{Sliczniuk2024}, the system is far from the saturation point, which indicates that the concentration gradient is controlled mainly by $c_s$. Consequently, the extraction kinetic term act as independent on $c_f$, which explain a uniform response in solid phase visible in Figure \ref{fig:Sensitivty_Gamma_CS}. The positive value of $\frac{dc_s}{d\Upsilon(0)}$ can be interpreted as a slower loss of the solute from particles and is observed when the extraction slows down. Because of the lower reaction rate, more solute remains in the particles initially, however, as time approaches infinity, the solute eventually diffuses out of the solid matrix. Consequently, it is expected to observed $\frac{dc_s}{d\Upsilon(0)}$ reaching a maximum and asymptotically diminish to zero.

@@

	Figure \ref{fig:Sensitivty_Gamma_CF} illustrates the sensitivity of the solute concentration in the fluid phase with respect to change into zero'th order term in the equation for $\Upsilon$. As previously discussed, the extraction rate slowed down due to increase of $\Upsilon(0)$ uniformly along the spatial domain. 
	Consequently, the solute delivered to the fluid phase decreases by an equal amount everywhere, causing a reduction in $c_f$, as indicated by the negative sign of  $\frac{dc_f}{d\Upsilon(0)}$. The spatial distribution of $\frac{dc_f}{d\Upsilon(0)}$ in the figure is non-uniform, reflecting the effects of advection. Due to mass conservation, the remaining solute reach the solid-fluid interface  more slowly hence it is expected to observe a little positive values of $\frac{dc_f}{d\Upsilon(0)}$ in the later stage of extraction. Eventually it is expected to see sensitivities reaching zero asymptotically.

@@

	Figure \ref{fig:Sensitivty_Gamma_Y}  shows the normalized sensitivity of the yield with respect to $\Upsilon(0)$. Due to slower extraction rate, the sensitivity curve is characterized by negative sign. The  downward trend of the sensitivity curve can be interpreted as less solute coming out of the extractor if compared with the reference case. The curve then reaches a negative minimum (also visible in Figure \ref{fig:Sensitivty_Gamma_Y}) before the remaining solute is eventually released to the fluid phase, causing the sensitivity to approach zero asymptotically. When the sensitivity function of a dynamic system approaches zero as time goes to infinity and remains at that value, it indicates that the system's long-term behaviour becomes independent of the parameter. The parameter has only influence for the system's transient dynamics, it has no effect on the steady-state solution. This means that after the system has settled, variations in that parameter do not affect the final output. In terms of SFE, it means that all the solute has been removed from the system.}

@@

	{\color{blue}The motivation for performing sensitivity analysis is to validate the model and evaluate how parameter variations affect its outputs By analysis the influence of parameters, potential modelling errors and unfeasible behaviours can be identified. This is particularly useful technique, when introducing an empirical correlation. In this work, the influence of the selected parameter from the empirical correlation for $\Upsilon$ was assessed. As expected, the analysis showed that increment of $\Upsilon(0)$ slows down the extraction rate, leading to negative values of $\frac{dc_s}{d\Upsilon(0)}$ and $\frac{dy}{d\Upsilon(0)}$, and alternating values of $\frac{dc_f}{d\Upsilon(0)}$. These findings align with the physical interpretation of $\Upsilon(0)$. The effect of the parameter change can be summarized by extreme and final values of the normalized sensitivity of the output (0.47 and 0, respectively).}

	The second analysis was to examine influence of operating conditions on the SFE model, particularly the pressure effect. Given that the results of the local sensitivity analysis depend on the operating point, the analysis was conducted at different pressures. For all the investigated operating conditions, it was observed that the pressure increment enhanced mass transfer, leading to a faster mass transfer of solute from the particles. Consequently, negative sensitivities can be observed in sensitivity diagrams for the solid phase. The corresponding response in the fluid phase is characterized by positive-valued sensitivities, indicating that higher quantity of solute is transported into the fluid phase. As a result, the extraction yield is improved and characterized by positive sensitivities. To compare the influence of various pressures, normalized sensitiviy coefficients were used, showing values between -$0.17$ and $0.6$. The overall effect of each parameter was assessed by integrating the yield sensitivity curves over extended simulation time (until all sensitivities approached zero). Table \ref{table:Cumulative_integral} presents the resulting cumulative integrals as well as initial and max values of the normalized sensitivity coefficients. The positive sign of the cumulative integrals confirm that increasing pressure improve the extraction yield under all investigated conditions. The relative magnitudes of these integrals indicate at which operating conditions are most affected by pressure variations are the most influential.

@@

	{\color{blue}The findings of this work are particularly valuable for industrial-scale supercritical extraction processes, where achieving both operational efficiency and product quality can be challenging. The local sensitivity analysis demonstrates that even small adjustments in operating parameters can yield disproportionately large effects on extraction rates, especially near the fluid’s critical point. By quantifying how parameter variations affect the model states, the high-efficiency operating setpoints can be identified. Moreover, linking local sensitivity analysis to numerical optimization establishes a systematic framework for industrial process design and scale-up. Rather than relying heavily on trial-and-error or resource-intensive pilot studies, the model-based approach can be used to pre-identify the region of interests. This approach reduces both the time and material costs needed to reach profitable extraction rates, moving industrial operations toward cost-effective production.}

@@

	% ===================================================
	% Bibliography
	% ===================================================
	%% Loading bibliography style file
	%\clearpage
	\newpage
	%\bibliographystyle{model1-num-names}
	\bibliographystyle{unsrtnat}
	\bibliography{mybibfile}
